### 2025-05-10 chapters/haiti/haiti.Rnw
Worked on appendices, table, arima2 chapter
Added appendix files, fixed citations, etc.

Regulation of biological populations is a fundamental topic in epidemiology, ecology, fisheries and agriculture.
Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
Quantitative models for these dynamic systems offer potential for designing effective control measures \cite{vandermeer13,he10}.
Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task \cite{king16}.
Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
What diagnostic tests and model variations can and should be considered in the course of the data analysis?
What are the possible trade-offs of increasing model complexity, such as the inclusion of interactions across spatial units?

This case study investigates the use of dynamic models and spatiotemporal data to inform public health policy in the context of the cholera outbreak in Haiti, which started in 2010.
Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
These modeling strategies were selected by Lee et a.~\cite{lee20} to represent the range of approaches used in the research community.
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
In addition, agent-based models were less widely used, the agent based model in \cite{lee20} being the only model of this class that was found in our literature review.
The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.

@@

# Plot reported cholera cases, by department. Creates Figure 1.
ggplot(dep_plot_df, aes(x = date, y = cases + 1)) +
  facet_wrap(~dep, nrow = 2, labeller = dep_labeller) +
  geom_line() +
  theme(axis.title.x = element_blank()) +
  ylab('Reported Cases') +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_x_date(
    date_labels = "'%y",
    breaks = seq.Date(
      from = as.Date("2011-01-01"),
      to = as.Date("2019-01-01"),
      by = '2 years'
    )
  )
@
\caption{\label{fig:CholeraData}
{\figTitle Weekly cholera cases.}
Weekly reported cholera cases in Haiti from October 2010 to January 2019 for each of the 10 administrative departments.
}
\end{figure}

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program.
The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{lee20,ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}.
In the Lee et al.~\cite{lee20} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti, estimating that a large number of cumulative cholera cases would be observed in the absence of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).
These forecasts are inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022 prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions, leading to recommendations for future studies.

The discrepancy between the model-based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.

Numerous guidelines have been proposed for using mechanistic models to inform policy, reviewed in \cite{behrend20}. Behrend et al.~\cite{behrend20} identify the importance of stakeholder engagement, transparency, reproducibility, uncertainty communication, and testable model outcomes. These and related principles are echoed by other influential articles \cite{saltelli20,donnelly18}. Additional literature emphasizes model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}. These guidelines often lack implementation specifics. As an example, \cite{lee20} largely adhere to the principles of \cite{behrend20}---though assessing the extent of stake-holder engagement is challenging---yet their projections are inconsistent with actual cholera incidence data from 2019 to 2022, demonstrating the limitations of current standards. We provide methodology for rigorous statistical calibration and evaluation of dynamic models (as advocated by \cite{saltelli19}), thereby expanding on the prevailing guidance. We specifically emphasize principles that prove essential in our case study. Complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \cite{li23}.

Our recommendations are presented in the context of a case study, with the goal of demonstrating how careful adherence to statistical principles may result in improved model fits.
We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
This includes a comprehensive tutorial on performing inference with Model~3 (\nameref{S_mod3cal}), a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.
We then use the improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
Finally, we conclude with a discussion of the results, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.

@@

Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}. More broadly, they have useful roles throughout biology, especially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}. In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}. The predictive skill of non-mechanistic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18,prosperi20}.

In this article, we refer to models that focus on learning relationships between variables in a dataset as {\it associative}, whereas models that incorporate a known scientific property of the system we call {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
Lucas et al.~\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.

The four mechanistic models of Lee et al.~\cite{lee20} were deliberately developed with limited coordination.
This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
Translations back to the original notation of Lee et al.~\cite{lee20} are given in the supplement (\nameref{S1_Table}).

@@

\begin{adjustwidth}{-2.25in}{0in}

@@

\caption{\label{tab:params} {\bf Model parameters.}}
\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.15\linewidth}t|p{0.17\linewidth}t|}

@@

Infection {\small (day)}

@@

Latency {\small (day)}

@@

  \multirow{3}{*}{Immunity {\small (yr)}}

@@

\multirow{2}{*}{Birth/death {\small (yr\textsuperscript{-1})}}

@@

  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)} {\small yr\textsuperscript{-1}}}

@@

  \Sexpr{myround(p3u["foi_add",1]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",2]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",3]*1e6,2)}, \allowbreak
  \Sexpr{myround(p3u["foi_add",4]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",5]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",6]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",7]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",8]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",9]*1e6,2)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
	) \allowbreak \times 10^{-6}$ {\small yr\textsuperscript{-1}}

@@

        \Sexpr{myround(p3u["betaB",1],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",2],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",3],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",4],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",5],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",6],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",7],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",8],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
        \Sexpr{myround(p3u["betaB",10],2)})$ {\small yr\textsuperscript{-1}}

@@

& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$ {\small yr\textsuperscript{-1}}} & \multirow{-3}{*}{\eqref{model2:lambda}} & &

@@

  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"]/52,3)} {\small wk\textsuperscript{-1}}

@@

Water survival {\footnotesize (wk)}

@@

\raggedright Process noise {\small (wk\textsuperscript{1/2})}

@@

  & (\nameref{S_meas})

@@

  & (\nameref{S_meas})

@@

  & \raggedright $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
  & \multirow{2}{*}{(\nameref{S_meas})}
  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(h2_fit$h2_params['v'], 3)}$}
  & \multirow{2}{*}{(\nameref{S_meas})}
  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
  & \multirow{2}{*}{(\nameref{S_meas})}

@@

  &  \multirow{2}{*}{(\nameref{S_init})}

@@

\begin{flushleft} References to the relevant equation are given in parentheses.
Parameters that were fixed and not calibrated using the data are indicated with $\fixed$; all fixed parameters values were chosen to match the fixed parameter values of \cite{lee20}.

@@

Confidence intervals for model parameters are given in the supplement (\nameref{S_CI}).
Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
\end{flushleft}
\end{adjustwidth}

@@

@@

While the complete model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.
Additional details about the numeric implementation of these models are provided in a supplemental text (\nameref{S1_Text}).
While each of the dynamic models considered in this manuscript can be fully described using the mathematical equations provided in the following section, diagrams of dynamic systems can be helpful to understand the equations.
For this reason, we provide flow chart diagrams for Models~1--3 in supplement figures (\nameref{S_Mod1}, \nameref{S_Mod2} and \nameref{S_Mod3}).

>>>>

Regulation of biological populations is a fundamental topic in epidemiology, ecology, fisheries and agriculture.
Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
Quantitative models for these dynamic systems offer potential for designing effective control measures \citep{vandermeer13,he10}.
Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task \citep{king16}.
Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
What diagnostic tests and model variations can and should be considered in the course of the data analysis?
What are the possible trade-offs of increasing model complexity, such as the inclusion of interactions across spatial units?

This case study investigates the use of dynamic models and spatiotemporal data to inform public health policy in the context of the cholera outbreak in Haiti, which started in 2010.
Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \citep{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
and 11 studies that used stochastic models \citep{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \citep{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
Here we focus on a multi-group modeling exercise by \citet{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
These modeling strategies were selected by \citet{lee20} to represent the range of approaches used in the research community.
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \citep{tracy18} are outside the scope of this article.
In addition, agent-based models were less widely used, the agent based model in \citet{lee20} being the only model of this class that was found in our literature review.
The data that were used in \citet{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.

@@

# Plot reported cholera cases, by department. Creates Figure 1.
ggplot(dep_plot_df, aes(x = date, y = cases + 1)) +
  facet_wrap(~dep, nrow = 2, labeller = dep_labeller) +
  geom_line() +
  theme(axis.title.x = element_blank()) +
  ylab('Reported Cases') +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_x_date(
    date_labels = "'%y",
    breaks = seq.Date(
      from = as.Date("2011-01-01"),
      to = as.Date("2019-01-01"),
      by = '2 years'
    )
  )
@
\caption[Weekly cholera cases]{\label{fig:CholeraData}
Weekly reported cholera cases in Haiti from October 2010 to January 2019 for each of the 10 administrative departments.
}
\end{figure}

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program.
The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \citep{lee20,ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}.
In the \citet{lee20} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti, estimating that a large number of cumulative cholera cases would be observed in the absence of additional vaccination efforts (Figure~3 and 4 of \citet{lee20}).
These forecasts are inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22}.
Though cholera has recently reemerged in Haiti \citep{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022 prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions, leading to recommendations for future studies.

The discrepancy between the model-based conclusions of \citet{lee20} and the prolonged absence of cholera in Haiti has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.

Numerous guidelines have been proposed for using mechanistic models to inform policy, reviewed by \citet{behrend20}, who identify the importance of stakeholder engagement, transparency, reproducibility, uncertainty communication, and testable model outcomes. 
These and related principles are echoed by other influential articles \citep{saltelli20,donnelly18}.
Additional literature emphasizes model calibration and evaluation techniques \citep{dahabreh17,egger17,penaloza15}. These guidelines often lack implementation specifics. As an example, \citet{lee20} largely adhere to the principles of \citet{behrend20}---though assessing the extent of stake-holder engagement is challenging---yet their projections are inconsistent with actual cholera incidence data from 2019 to 2022, demonstrating the limitations of current standards. We provide methodology for rigorous statistical calibration and evaluation of dynamic models (as advocated by \citet{saltelli19}), thereby expanding on the prevailing guidance. We specifically emphasize principles that prove essential in our case study. Complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \citet{li23}.

Our recommendations are presented in the context of a case study, with the goal of demonstrating how careful adherence to statistical principles may result in improved model fits.
We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
This includes a comprehensive tutorial on performing inference with Model~3 (Appendix~\ref{sec:appendix_haiti_mod3Cal}), a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.
We then use the improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by \citet{lee20}.
Finally, we conclude with a discussion of the results, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.

@@

Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \citep{lofgren14,mccabe21}. More broadly, they have useful roles throughout biology, especially when combined with statistical methods that properly account for stochasticity and nonlinearity \citep{may04}. In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \citep{lau22,baker18}. The predictive skill of non-mechanistic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \citep{baker18,prosperi20}.

In this article, we refer to models that focus on learning relationships between variables in a dataset as {\it associative}, whereas models that incorporate a known scientific property of the system we call {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \citep{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.

The four mechanistic models of \citet{lee20} were deliberately developed with limited coordination.
This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
Translations back to the original notation of \citet{lee20} are given in Table~\ref{tab:translate}.

@@

% \begin{adjustwidth}{-2.25in}{0in}

@@

\footnotesize
\begin{tabular}{|p{0.165\linewidth}|p{0.16\linewidth}t|p{0.17\linewidth}t|p{0.165\linewidth}t|}

@@

Infection {(day)}

@@

Latency {(day)}

@@

  \multirow{3}{*}{Immunity {(yr)}}

@@

\multirow{2}{*}{Birth/death {(yr\textsuperscript{-1})}}

@@

  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)} \allowbreak {\tiny yr\textsuperscript{-1}}}

@@

  \Sexpr{myround(p3u["foi_add",1]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",2]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",3]*1e6,1)}, \allowbreak
  \Sexpr{myround(p3u["foi_add",4]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",5]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",6]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",7]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",8]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",9]*1e6,1)}, \allowbreak
	\Sexpr{myround(p3u["foi_add",10]*1e6,1)}
	) \allowbreak \times 10^{-6}$ {yr\textsuperscript{-1}}

@@

        \Sexpr{myround(p3u["betaB",1],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",2],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",3],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",4],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",5],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",6],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",7],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",8],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",9],1)}, \allowbreak
        \Sexpr{myround(p3u["betaB",10],1)})$ {yr\textsuperscript{-1}}

@@

& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$ {yr\textsuperscript{-1}}} & \multirow{-3}{*}{\eqref{model2:lambda}} & &

@@

  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"]/52,3)} {wk\textsuperscript{-1}}

@@

\raggedright Water survival (wk)

@@

\raggedright Process noise {(wk\textsuperscript{1/2})}

@@

  & \eqref{model1:meas}

@@

  & \eqref{model2:meas}

@@

  & \raggedright $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],1)}, \Sexpr{myround(p1["tau_end"],1)})$
  & \multirow{2}{*}{\eqref{model1:meas}}
  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(h2_fit$h2_params['v'], 1)}$}
  & \multirow{2}{*}{\eqref{model2:meas}}
  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],1)}$}
  & \multirow{2}{*}{\eqref{model3:meas}}

@@

@@

\caption[Parameter estimates of the cholera models.]{\label{tab:params}
References to the relevant equation are given in parentheses.
Parameters that were fixed and not calibrated using the data are indicated with $\fixed$; all fixed parameters values were chosen to match the fixed parameter values of \citet{lee20}.

@@

Confidence intervals for model parameters are given in Appendix~\ref{sec:appendix_haiti_ci}.
Translations back into the notation of \citet{lee20} are given in Table~\ref{tab:translate}.
}
% \begin{flushleft}
% \end{flushleft}
% \end{adjustwidth}

@@

\begin{table}
  \begin{center}
  \begin{tabular}{|c|c|c|c|c|}\hline
    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
     & Notation & 1 & 2 & 3 \\
    \hline
    \hline
    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $1-\kappa$ & $red_\beta$ & --- \\\hline
    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
  \end{tabular}
  \end{center}
  \caption[Parameter translations to original model definitions]{\label{tab:translate}Translations between our common notation and notation used by Lee et al (2020).}
\end{table}

@@

While the complete model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.
Additional details about the numeric implementation of these models are provided in Appendix~\ref{sec:appendix_haiti_details}.
While each of the dynamic models considered in this manuscript can be fully described using the mathematical equations provided in the following section, diagrams of dynamic systems can be helpful to understand the equations.
For this reason, we provide flow chart diagrams for Models~1--3 in Figures~\ref{fig:flow_diagram}--\ref{fig:flow_diagram3}.

\subsubsection*{Model~1}
\label{sec:model1}
The latent state vector $\bm{X}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$ at time $t$.
Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs. Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, separating individuals into compartments with distinct dynamics based on vaccination status.
The force of infection is
\begin{equation}
\label{model1:lambda}
\lambda(t) = \Big(\sum_{\vaccCounter=0}^{\vaccClass}  I_{\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^{\vaccClass} A_{\vaccCounter}(t) \Big)^\nu \frac{d\Gamma(t)}{dt} \transmission(t)/N,
\end{equation}
where $\transmission(t)$ is a periodic cubic spline representation of seasonality, given in terms of a B-spline basis $\{ s_j(t), j\in \seq{1}{6}\}$ and parameters $\transmission_{1:6}$ as
\begin{equation}
\label{model1:beta}
\transmission(t) = \bar{\transmission} \exp\Big(\sum_{j=1}^6 \transmission_j s_j(t)\Big),
\end{equation}
where $\bar{\transmission} = 1$ (wk)\textsuperscript{-1} is a dimensionality constant.
The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
\citet{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \citep{breto11} which has been found to improve the statistical fit of disease transmission models \citep{stocks20,he10}.
